Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-08-10 paper.tex
forgot to save
more

transitivity : {A : Set} {x y z : A} → (x ≡ y) →
               (y ≡ z) → (x ≡ z)
\end{code}
\end{minipage}
\end{center}

@@

reverse : {A : Set} → List A → List A
append-left-unit : {A : Set} (xs : List A) → Nil ++ xs ≡ xs
append-right-unit : {A : Set} (xs : List A) → xs ++ Nil ≡ xs
append-associativity : {A : Set} (xs ys zs : List A) →
                       xs ++ (ys ++ zs) ≡ (xs ++ ys) ++ zs
append-distributivity : {A : Set} (xs ys : List A) →
                        reverse (xs ++ ys) ≡ (reverse ys) ++ (reverse xs) 
double-reverse : {A : Set} (xs : List A) → reverse (reverse xs) ≡ xs
\end{code}
\end{minipage}
\end{center}

\subsection*{Extension to Homotopy Type Theory (maybe)}
Earlier, we briefly discussed how types can be viewed as shapes, with their
respective terms consequently viewable as points contained with shapes. This
concept is important in understanding another way to view the concepts of type
theory: {\em homotopy type theory} (HoTT).

In fact, a lot of what HoTT seeks to do stems directly from a single concept in
regular type theory: the induction principle of identity types! This is why we
briefly discussed identity types having {\em another}, less general induction
principle. This less general induction principle is known as {\tt K} and is
defined as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
K : {A : Set} (C : {x : A} → x ≡ x → Set) →
    (∀ (x : A) → C (refl x)) →
    {x : A} (p : x ≡ x) → C p
K C c (refl x) = c x
\end{code}
\end{minipage}
\end{center}
{\tt K} is similar to {\tt J} but asserts that the only proof of equality is
{\tt refl} (i.e., the only equality is reflexive equality). In fact, we can
prove this using {\tt K}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
UIP≡ : {A : Set} {x : A} (p : x ≡ x) → p ≡ (refl x)
UIP≡ = 
  K (λ {x} x=x → x=x ≡ (refl x))
    (λ x → refl (refl x))
\end{code}
\end{minipage}
\end{center}
As we mentioned earlier, we cannot prove UIP for identity types using {\tt J},
since the identity type passed to {\tt J} is parameterized over two possibly
different terms, while {\tt K} only applies to reflexive proofs of equality.
In some cases, {\tt K} might be all that one would need to do a given number
of proofs concerning identity types, however, it limits the overall application
and expressivity of identity types.
\subsection*{Conclusion and Acknowledgments}

\end{document}

>>>>

ap : {A : Set} {B : Set} (f : A → B) →
     {x y : A} → (x ≡ y) → (f x ≡ f y)
transitivity : {A : Set} {x y z : A} → (x ≡ y) →
               (y ≡ z) → (x ≡ z)
\end{code}
\end{minipage}
\end{center}

@@

rev : {A : Set} → List A → List A
append-left-unit : {A : Set} (xs : List A) → Nil ++ xs ≡ xs
append-right-unit : {A : Set} (xs : List A) → xs ++ Nil ≡ xs
append-associativity : {A : Set} (xs ys zs : List A) →
                       xs ++ (ys ++ zs) ≡ (xs ++ ys) ++ zs
append-distributivity : {A : Set} (xs ys : List A) →
                        rev (xs ++ ys) ≡ (rev ys) ++ (rev xs) 
double-rev : {A : Set} (xs : List A) → rev (rev xs) ≡ xs
\end{code}
\end{minipage}
\end{center}

\subsection*{A little more on Identity Types}
Earlier, we briefly discussed how types can be viewed as shapes, with their
respective terms viewable as points contained within them. This view of types
is important in understanding another way to apply type theory:
{\em homotopy type theory} (HoTT).

In fact, a lot of what HoTT seeks to do stems directly from a single concept in
regular type theory: the induction principle of identity types! This is why we
briefly discussed identity types having {\em another}, less general induction
principle. This less general induction principle is known as {\tt K} and is
defined as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
K : {A : Set} (C : {x : A} → x ≡ x → Set) →
    (∀ (x : A) → C (refl x)) →
    {x : A} (p : x ≡ x) → C p
K C c (refl x) = c x
\end{code}
\end{minipage}
\end{center}
{\tt K} is similar to {\tt J} but asserts that the only proof of equality is
{\tt refl} (i.e., the only equality is reflexive equality). In fact, we can
this concept using {\tt K}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
UIP≡ : {A : Set} {x : A} (p : x ≡ x) → p ≡ (refl x)
UIP≡ = 
  K (λ {x} x=x → x=x ≡ (refl x))
    (λ x → refl (refl x))
\end{code}
\end{minipage}
\end{center}
As we mentioned earlier, we cannot prove {\tt UIP≡} using {\tt J}, since the
identity type passed to {\tt J} is parameterized over two possibly different
terms, while {\tt K} only applies to reflexive proofs of equality. Another way
to think of this is that {\tt J} applies to all two terms that are proven equal
{\em in some way}, while {\tt K} can only be used for terms that either
reducible to each other or are, in some way, the same. In some cases, {\tt K}
might be all that one would need to do a given number of proofs concerning
identity types, however, it limits the overall application and expressivity of
identity types.

In HoTT, identity types are interpretted rather differently than in regular type
theory. In HoTT, identity types correspond to {\em topological paths} that
connect two elements. Here, we say {\em elements} because these paths can
connect points, paths, paths of paths, etc. Another way to say this is that HoTT
has the ability to reason about the {\em equality of equalities}. This is the
reason for complexity of {\tt J} and is also the reason why {\tt K} is not
applicable to HoTT.

To end our brief discussion on identity types, it helps to further to solidify
the concept of how identity types can be interpretted as topological paths in
HoTT. For simplicity, we show how all the methods in which we can interact with
an identity type correspond to a way one could think of interacting with a path.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
Identity Type → Path

J → path induction
symmetry → path inverse
ap → function application on paths
congruence → proposition interaction on paths
transitiviy → path concatenation
\end{code}
\end{minipage}
\end{center}

\subsection*{Conclusion and Acknowledgments}
At its core, the main purpose of type theory is to propose a new way of doing
mathematics. The main benefit of choosing to do so stems from the close
connection between types, constructive logic and the field of computer science.
Thus, through simple act of type checking, which was made possible due in part
to type theory being founded on the STLC, proofs and programs can now be
verified and corrected with the help of a computer equipped with a proof
assistant, which has given both mathematics and computer science the powerful
concept of {\em proof verification}.

I would like to thank Daniel P. Friedman for guiding me as I explored the world
of types and type theory. 
\end{document}
